---
title: "lab10"
format: pdf
editor: visual
---

# 1. Importing candy data

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

> **Q1**. How many different candy types are in this dataset?

`r nrow(candy)`

> **Q2**. How many fruity candy types are in the dataset?

`r sum(candy["fruity"] == 1)`

> **Q** What are these fruity candy

```{r}
rownames(candy[candy$fruity == 1, ])
```

# 2. What is your favorite candy?

How often does my favorite candy win?

```{r}
candy["Twix", ]$winpercent
```

```{r}
candy["Twix",]$winpercent
```

```{r}
# install.packages("skimr")
# skimr::skim(candy) # if you only want to use this one funciton from this pacakge
library("skimr")
skim(candy)
```

> **Q6**. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes. Winpercent column has mean that's a few times bigger than the means of other columns.

> **Q7**. What do you think a zero and one represent for the `candy$chocolate` column?

Zero means the candy is not classified as containing chocolate.

> **Q8**. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy) + 
  aes(x = winpercent) +
  geom_histogram(bins = 39)
```

> **Q9**. Is the distribution of winpercent values symmetrical?

No

> **Q10**. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

Above

> **Q11**. On average is chocolate candy higher or lower ranked than fruit candy?

I need to "subset" (a.k.a. "select", "filter") to just chocolate candy, get their winpercent values, and then calculate the means of these.

```{r}
# Filter/select/subset to just chocolate rows
chocolate.candy <- candy[as.logical(candy$chocolate), ]

# Get their winpercent values
chocolate.winpercent <- chocolate.candy$winpercent

# Calculate their mean winpercent value
mean(chocolate.winpercent)
```

```{r}
mean(candy[candy$chocolate == 1, ]$winpercent)
mean(candy[candy$fruity == 1, ]$winpercent)
```

Higher

> **Q12**. Is this difference statistically significant?

```{r}
chocolate <- candy[candy$chocolate == 1, ]$winpercent
fruity <- candy[candy$fruity == 1, ]$winpercent
t_test <- t.test(chocolate, fruity, alternative = c("greater")) # is chocolate's mean statistically significantly greater than fruity's mean?
```

P-value `r t_test$p.value` \< 0.05, so statistically significant

# 3. Overall Candy Rankings

`order()` returns the "indices" of the input that would result in it being sorted.

> **Q13**. What are the five least liked candy types in this set?

```{r}
# sort(candy$winpercent, decreasing=FALSE)
rownames(candy[order(candy$winpercent, decreasing=FALSE), ][1:5, ])
```

> **Q14**. What are the top 5 all time favorite candy types out of this set?

```{r}
rownames(candy[order(candy$winpercent, decreasing=TRUE), ][1:5, ])
```

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_bar(stat = "identity")
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) + 
  geom_bar(stat = "identity")
```

### Time to add some useful color

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```

> **Q17**. What is the worst ranked chocolate candy?

Sixlets

> **Q18**. What is the best ranked fruity candy?

Starburst

# 4. Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
my_cols[as.logical(candy$fruity)] == "red"
ggplot(candy) +
  aes(winpercent, pricepercent) +
  geom_point(col=my_cols)
```

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel()
```

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> **Q19**. Which candy type is the highest ranked in terms of `winpercent` for the least money - i.e. offers the most bang for your buck?

Reese's minatures

> **Q20**. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip

# 5 Exploring the correlation structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

> **Q22**. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

fruity and chocolate

> **Q23**. Similarly, what two variables are most positively correlated?

winpercent and chocolate

# 6. Principal Component Analysis

Let's apply PCA using the `prcom()` function to our candy dataset remembering to set the `scale=TRUE/FALSE` argument.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
pca_unscaled <- prcomp(candy, scale=FALSE)
summary(pca_unscaled)
```

Now we can plot our main PCA score plot of PC1 vs PC2.

```{r}
plot(pca$x[, 1], pca$x[, 2])
```

```{r}
plot(pca_unscaled$x[, 1], pca_unscaled$x[, 2])
```

We can change the plotting character and add some color:

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
pc <- as.data.frame(pca$x)

ggplot(pc) + 
  aes(x=PC1, y=PC2, label=rownames(pc)) +
  geom_point(col=my_cols) + 
  geom_text_repel(max.overlaps = 5)
```

> **Q24**. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

Fruity. PC1 is strongly correlated with fruity variable
