---
title: "Class 08: Machine learning"
format: pdf
editor: visual
---

```{r}
wisc.data <- read.csv("WisconsinCancer.csv", row.names=1)
head(wisc.data)
```

> Q How many patient samples are in this dataset?

```{r}
nrow(wisc.data)
```

There're are `r nrow(wisc.data)` patients in this dataset

```{r}
table(wisc.data$diagnosis)
```

```{r}
wisc <- wisc.data[,-1]
```

```{r}
diagnosis <- as.factor(wisc.data$diagnosis)
# diagnosis
```

```{r}
ncol(wisc)
```

There're `r ncol(wisc)` variables in this dataset\

## Principle Component Analysis

```{r}
round(colMeans(wisc))
```

```{r}
pca <- prcomp(wisc, scale=TRUE)
summary(pca)
```

```{r}
attributes(pca)
```

> Q Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
plot(pca$x[,1], pca$x[,2], col=diagnosis)
```

```{r}
library(ggplot2)

x <- as.data.frame(pca$x)

ggplot(x) +
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```

> Q How much variance is captured in the top three PCs?

They capture 72.64%

> **Q** For the first principal component, what is the component of the loading vector (i.e. `wisc.pr$rotation[,1]`) for the feature `concave.points_mean`?

```{r}
pca$rotation["concave.points_mean", 1]
```

## Combine PCA results with clustering

We can use our new PCA variables (i.e. the scores along the PCs contained in `pca$x`) as input for other methods such as clustering.

```{r}
d <- dist(pca$x[, 1:3]) # Hclust needs a disstance matrix as input
hc <- hclust(d, method="ward.D2")
plot(hc)
```

To generate our cluster membership vector, we can use the `cutree()` function and specify a height (`h`) or number of groups (`k`).

```{r}
grps <- cutree(hc, h=80)
table(grps)
```

```{r}
table(diagnosis)
```

```{r}
table(diagnosis, grps)
```

```{r}
library(ggplot2)

ggplot(x) +
  aes(PC1, PC2) + 
  geom_point(col=grps)
```

> Q What is the specificity and sensitivity of our current results?

Sepc = TN/(TN + FP) = 333/(333 + 33) = `r 333/(333 + 33)`

Sens = TP/(TP+FN) = 179/(179+24) = `r 179/(179+24)`

## Prediction

> Q Which of these new patients should we prioritize for follow up based on your results?

Prioritize patient 2, because it's malignant.
